BOLD Parameter Estimation Using Particle Filters

### 2010-07-03 doc/thesis/methods.tex
mostly done with methods for real this time

In normal statistical parametric mapping, a gaussian smoothing
filter is applied across the image as discussed in \autoref{sec:RFT}.
After this, detrending is performed discussed by \autoref{sec:Detrend}.
Recall that FMRI signal levels are unit-less and even though detrending isn't
always necessary, at the very least it is necessary to convert
the data into \% difference from the baseline. Changing to 
\% difference removes no real data
from signal since it merely subtracting then dividing by a constant. This
is the signal that was input into the delta based particle filter.
Of course there are much more advanced ways of performing this task.
The generally accepted standard is actually to use a high pass filter, although the
cutoff frequency is application dependent and often applied haphazardly.
The high pass filter thus removes the DC component of the signal, and 
some amount of the so called "drift". The problem with this method is that it is
not adaptive to the input. Huge variations in drift frequencies can exist 
in a single time-series. Thus, a single cutoff frequency could miss a significant
drift component, or it could remove \emph{actual} signal, if the cutoff frequency is
set too high. This is why, as I mentioned in the De-trending section, a spline
based detrending method will generally give better results.

For simulated and real images (tests with multiple time-series), tests were 
also run with and without Gaussian filtering with sigma of %not sure todo
were run, since it is standard
practice to apply a Gaussian spatial filter to the images at each timestep. Obviously
a spatial filter such as Gaussian filtering increased SNR but can also lead to less
precision in the output maps.

\section{Simulation}
We performed two types of simulations. First, we simulated a single BOLD time-series based
on a random chosen set of model parameters. This process was relatively straight forward
given the state-space equations for the BOLD signal. After a "true" signal was generated,
we then added a carrier level, since BOLD is typically measured as a \% difference from the
base level. Finally, we added Gaussian noise, and a Weiner Process to the clean signal. The
variance of the Gaussian noise may be expressed in terms of the desired noise SNR, $R$ as:
\begin{equation}
var(y_{noisy}) = var(y) / R
\end{equation}
Since SNR doesn't have quite the same meaning for a Wiener process based noise, the variance 
of the Gaussian steps was set to be:
\begin{equation}
var(y_{noisy}) = var(y) / (4R)
\end{equation}
Once this noisy simulated time series was generated, the exact same particle filter algorithm
that would later be run on full sized images, was run on this single voxel image. We ran
a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

Second we used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by creating a random image, smoothing it with a large Gaussian kernel, then thresholding
the results. Finally connected regions were each given a set of parameters from a finite
list of randomly chosen parameter sets. The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Time-series of activation level
was generated for each set of parameters, then activation levels were fed into POSSUM's 
function for generating frequency domain data. The patche for POSSUM will be made available.
For each time-series in the simulate FMRI image, the final \emph{static} parameters are saved
into a parameter map. This parameter map may then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters may also be 
generated to test the functional difference between the two maps. This would give an absolute 
quantitative difference between the two parameter sets irrespective to parameter slopiness.
So for instance, if $V_0$ is halved, $\epsilon$ doubling may very well give a similar result.
In this case the \% difference between the parameters will be large in each case, but the functional
difference between the parameters will not be great. This is obviously a bad situation, which
is why we wanted to test for it.

@@

The final result from calculating parameters with the real data was similar to that
from the results from the POSSUM simulated data. The difference being that there was
no ground truth the check it with.

>>>>

In normal statistical parametric mapping, a gaussian smoothing
filter is applied across the image as discussed in \autoref{sec:RFT}.
After this, detrending is performed discussed by \autoref{sec:Detrend}.
Recall that FMRI signal levels are unit-less and even though detrending isn't
always necessary, at the very least the data must be converted 
into \% difference from the baseline. Changing to 
\% difference removes no real information
from signal. This is the signal that was input into the delta based 
particle filter. Of course, most of the time analysis is performed on the
direct signal; which mandates the removal of low frequency drift.
The generally accepted method is to use a high pass filter, although the
cutoff frequency is application dependent and often applied haphazardly.

%For simulated and real images (tests with multiple time-series), tests were 
%also run with and without Gaussian filtering with sigma of %not sure todo
%were run, since it is standard
%practice to apply a Gaussian spatial filter to the images at each timestep. Obviously
%a spatial filter such as Gaussian filtering increased SNR but can also lead to less
%precision in the output maps.

\section{Simulation}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a carrier level, since BOLD is typically measured as a \% difference from the
base level. I tested the particle filter with two types of noise; independent and identically
distributed gaussian noise and weiner drift. For both I tested with various standard deviations.

Once this noisy simulated time series was generated, the exact same particle filter algorithm
that would later be run on full sized images, was run on this single voxel image. I ran
a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

Second I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this should not have an effect on the parameters (other than perhaps $\epsilon$ and $V_0$). 
For each time-series in the simulated FMRI image, the final \emph{static} parameters were saved
into a parameter map. This parameter map may then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters may also be 
generated to test the difference in activation levels between the real parameters and the
estimated ones. Since it is clear that the parameters are not fully orthogonal 
(\cite{Deneux2006}), its possible that two sets of parameters are functionally equivalent,
but have different parameters. This way, an absolute 
quantitative difference between the two parameter sets may be found.

@@

Before performing tests on a full image, I tested the results of the particle filter
on regions deemed active and non-active by statistical parametric mapping.
